---
title: "Preparation of input to run PANDA on Python"
author: "Victor Hugo Calegari de Toledo"
date: "February 21, 2020"
output:
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: yes
    df_print: paged
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This document describes how we create inputs of our fetal brain data for PANDA.

```{r echo = FALSE, message = FALSE, warning = FALSE}
# Packages needed:
#remotes::install_github("netZoo/netZooR", build_vignettes = TRUE)
library(dplyr)
library(data.table)
#library(pandaR)
library(parallel)
library(foreach)
library(org.Hs.eg.db)
library(gtools)
library(doParallel)
library(clusterProfiler)
library(reshape)
library(netZooR)
library(ggplot2)
library(tidyverse)
library(visNetwork)
library(fgsea)
options(stringsAsFactors = FALSE)
options(bitmapType = 'cairo')
```

## PANDA input processing

The motif file and the correspondence of their IDs to Gene Symbols were downloaded from Kimberly Glass PANDA [page]<https://drive.google.com/file/d/1MlPiAmOxu4mTd4nW9zMygzK1U9F8XB0M/view?usp=sharing> (23.10.2019).
The PPI file was downloaded from StringDB [(version 11.5)<https://stringdb-static.org/download/protein.links.v11.5/9606.protein.links.v11.5.txt.gz>.
The expression and ppi data were converted to Gene Symbol, which is used in PANDA's input files.
 
### Annotation conversion:
Initial number of interactions:
```{r echo = FALSE, message = FALSE, warning = FALSE}
#First we will load the motif files and convert the motif ID to their correspondent gene symbol:

## Preparing the data before running PANDA:
  # Starting with the motif data
  setwd("~/projects/sex_differences_fetal_brain/netZoo/panda/_m")
  fread("~/projects/sex_differences_fetal_brain/netZoo/panda/_h/hg38_refseq_-750_250_1e-4", sep='\t') -> motif
  read.delim("~/projects/sex_differences_fetal_brain/netZoo/panda/_h/Homo_sapiens_motifinfo.txt", sep='\t') -> anot
  gsub("_.*","", anot$MotifID) -> anot$MotifID
  gsub("^\\(|\\)_.*$", "", anot$TF_Info) -> anot$TF_Info

  # Convert them to their gene symbol according to annotation:
  unique(motif$V1) -> mt
  mt[(mt %in% anot$MotifID)] -> mt #all of them are present
  motif[motif$V1 %in% mt,] -> motif #all of them are present
  '%!in%' <- function(x,y)!('%in%'(x,y))
  anot[anot$MotifID %!in% mt,1] #"M0949" "M1950" "M1970"
  anot[anot$MotifID %in% mt,] -> anot #3 of them were not present
  motif %>% inner_join(anot, by=c("V1" = "MotifID")) -> motif
  motif[,c(4,2,3)] -> motif
  colnames(motif) = c("tf","target", "weight")
  nrow(motif)
```
Initial number of genes:
```{r echo = FALSE, message = FALSE, warning = FALSE}
length(unique(motif$target))

```
Checking for missing annotation:
```{r echo = FALSE, message = FALSE, warning = FALSE}

  # Manually imputing the missing gene symbols:
  unique(motif$tf) # checking each of the motifs manually, the missing are:
  # ENSG00000250542 (19) > no alias
  # ENSG00000235187 (49) > no annotation at all
  # AC226150.2 (67) > no alias
  # ENSG00000229544 (90) > NKX1-2
  motif$tf[motif$tf == "ENSG00000229544"] = "NKX1-2"
  
  #anot2 = bitr(motif$target,fromType="SYMBOL",toType="ENSEMBL",OrgDb="org.Hs.eg.db")
  #motif %>% inner_join(anot2,by=c("target" = "SYMBOL")) -> motif_panda
  #nrow(motif_panda)
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
  # We didn't need to convert to ENSEMBL
  #mia = motif[!motif$target %in% motif_panda$target,]
  #anot3 = bitr(mia$target,fromType="ALIAS",toType="ENSEMBL",OrgDb="org.Hs.eg.db")
  #mia %>% inner_join(anot3,by=c("target" = "ALIAS")) -> motif_panda2
  #motif_panda <- rbind(motif_panda,motif_panda2)
  #mia2 = motif[!motif$target %in% motif_panda$target,]
  # Most of what is missing is not properly curated, like the ^LOCs, or are pseudogenes
```
  Check them number of canonical interactions:
```{r echo = FALSE, message = FALSE, warning = FALSE}

# Transform the target column in the motif file in Ensembl Gene annotation:
   #final_motifs = motif_panda[,c(1,4,3)]
   #colnames(final_motifs) = c("tf","target","canonical")
   table(motif$weight) #only canonical interactions
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
  fread("~/projects/sex_differences_fetal_brain/netZoo/panda/_h/corrected_exp.txt", sep='\t', header = T) -> exp_ensembl
  gene_symbols = bitr(expr$Genes,fromType="ENSEMBL",toType="SYMBOL",OrgDb="org.Hs.eg.db")
  expr %>% inner_join(gene_symbols,by=c("Genes" = "ENSEMBL")) -> exp
  exp$Genes <- exp$SYMBOL
  exp <- exp[,1:120]
  
  # 18137 genes
  # 46 duplicated
  exp <- exp[,lapply(.SD, mean), exp$Genes]
  # 18091 genes
  
 colnames(exp)[1] <- "Genes"
  
  # ## FIRST: filter the motif by the expression files
  motif_hit = motif[(motif$target %in% exp$Genes),] 
  
  # New number of interactions: 3405257
  # Number of lost interactions: 3995520

  # ## SECOND: filter the expression by the motif files:
  exp = exp[(exp$Genes %in% motif$target),]
  
  # New number of genes: 15150

```

Creating the files to run PANDA in Python in the cluster:
```{r echo = FALSE, message = FALSE, warning = FALSE, results = FALSE}
  
  # Now we will load the TFs interaction network and the aliases from other annotations:
  ## Download the protein links from the stringDB:
  string = fread('~/projects/sex_differences_fetal_brain/netZoo/panda/_h/9606.protein.links.v11.5.txt.gz.inproc', sep=' ')
  string = string[string$combined_score >= 700,]
  anotstring = fread('~/projects/sex_differences_fetal_brain/netZoo/panda/_h/9606.protein.aliases.v11.5.txt', sep='\t', header=TRUE)
  colnames(anotstring)[1] <- "string_protein_id"
  anotstring$string_protein_id = gsub("9606.","",anotstring$string_protein_id)
  string$protein1 = gsub("9606.","",string$protein1)
  string$protein2 = gsub("9606.","",string$protein2)
  
  # Convert the protein ENSEMBL IDs to gene symbol:
  gene_symbols2 = bitr(string$protein1,fromType="ENSEMBLPROT",toType="SYMBOL",OrgDb="org.Hs.eg.db")
  length(unique(string$protein1)) # 16814
  string %>% inner_join(gene_symbols2,by=c("protein1" = "ENSEMBLPROT")) -> ppi
  
  gene_symbols3 = bitr(string$protein2,fromType="ENSEMBLPROT",toType="SYMBOL",OrgDb="org.Hs.eg.db")
  ppi <- ppi[,c(4,2,3)]
  ppi %>% inner_join(gene_symbols3,by=c("protein2" = "ENSEMBLPROT")) -> ppi
  ppi <- ppi[,c(1,4,3)]
  colnames(ppi) <- c("protein1","protein2","score")
```  
  
  Formating the files to run in Python:
  
```{r echo = FALSE, message = FALSE, warning = FALSE, results = FALSE}
    # Reading metadata
    pheno <- read.delim("~/projects/sex_differences_fetal_brain/panda/_h/phenotype.txt",check.names = F,header = T)
    pheno$Samples <- as.character(pheno$Samples)
    row.names(pheno) <- pheno$Samples
    pheno[,c(2,3,5)] <- lapply(pheno[,c(2,3,5)], factor)
  
  # Saving the PANDA output just in case:
  save(exp,motif_hit,ppi,file="~/projects/sex_differences_fetal_brain/netZoo/panda/_m/panda_input.rda")
  
  ##### Expression files:
  exp <- as.data.frame(exp)
  female_samples = subset(pheno$Samples,pheno$Sex == 0)
  female_exp = exp %>% select(1,which(colnames(exp) %in% female_samples))
  male_samples = subset(pheno$Samples,pheno$Sex == 1)
  male_exp = exp %>% select(1,which(colnames(exp) %in% male_samples))
  
  
  
  write.table(male_exp, file = "~/projects/sex_differences_fetal_brain/netZoo/panda/_m/male_exp.txt", quote = FALSE, row.names = FALSE, col.names = FALSE, sep = "\t")

  write.table(female_exp, file = "~/projects/sex_differences_fetal_brain/netZoo/panda/_m/female_exp.txt", quote = FALSE, row.names = FALSE, col.names = FALSE, sep = "\t")
  
  ##### PPI files:
  write.table(ppi,file = "~/projects/sex_differences_fetal_brain/netZoo/panda/_m/ppi.txt", quote = FALSE, row.names = FALSE, col.names = FALSE, sep = "\t")
  
  ##### Motif files:
  write.table(motif_hit,file = "~/projects/sex_differences_fetal_brain/netZoo/panda/_m/motif.txt", quote = FALSE, row.names = FALSE, col.names = FALSE, sep = "\t")
```

```{r echo = FALSE, message = FALSE, warning = FALSE, results = FALSE}
sessionInfo()
```